---
title: "Basic of Explanatory Data Analysis"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: 'show'
#date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo =T, warning=F, message=F, comment='')
suppressWarnings(
  suppressMessages(
    suppressPackageStartupMessages({
      library(stats)
      library(MASS)
      library(tidyverse)
      library(magrittr)
      library(effsize)
      library(corrplot)
      library(ppcor)
      library(gridExtra)
      library(palmerpenguins)
    })
  )
)
options(scipen=100,digits=3)
```

# random variables, values from repeated observation
$x_{i=1-n}$

- continuous
    - x is real
- discreate
    - x is integer
        - count x>=0
        - binary x{0,1}


# table data
$X(x_{i=1-n,j=1-p})$

- cross sectional data
    - $x_{,j}$ have distribution
    - $x_{i,j}$ and $x_{i',j}$ are independent
- panel data
    - $x_{,j}$ have distribution
    - $x_{i,j}$ and $x_{i',j}$ are independent
    - $x_{i,j}$ and $x_{i,j'}$ have relation(paired, class)
- time series data
    - $x_{,j}$ have (time)order, auto correlation, trend, seasonality


## utility functions
```{r}
#skewness of vector
skew=function(x){
  m=mean(x,na.rm=T)
  mean((x-m)**3,na.rm=T)/(mean((x-m)**2,na.rm=T))**1.5
}

#kurtosis of vector
kurt=function(x){
  m=mean(x,na.rm=T)
  mean((x-m)**4,na.rm=T)/(mean((x-m)**2,na.rm=T))**2
}

#standardize[0,1] of vector
std0to1=function(x){
  (x-min(x,na.rm=T))/(max(x,na.rm=T)-min(x,na.rm=T))
}

#summarize variables by arsenal
smArs=function(tb){
  arsenal::tableby(~.,tb) |> 
    summary(text=T) |>
    kable('simple')
}

#rows have missing data
miss=function(tb){
  tb[!complete.cases(tb),]
}

#eject outlier of a vector
outlier=function(x){
  #boxplot(scale(x))
  q=fivenum(x)
    r=q[4]-q[2]
    l=q[2]-r*1.5
    u=q[4]+r*1.5
    x=x[x>l & x<u]
  #boxplot(x)
  x
}

#eject outlier of all numeric variables
outliers=function(tb){
  par(mfrow=c(1,2))
  v=select(tb,is.numeric) |> colnames()
  #boxplot(scale(tb[v]))
  
  q=tb[v] |> sapply(fivenum)
  for(i in v){
    r=q[4,i]-q[2,i]
    l=q[2,i]-r*1.5
    u=q[4,i]+r*1.5
    tb=tb[(tb[i]>l & tb[i]<u),]
  }
  #boxplot(scale(tb[v]))
  tb
}

#see distribution of all numeric variables
dists=function(tb){
  options(digits=2)
  par(mfrow=c(2,2))
  c=select(tb,is.numeric) |> colnames()
  tb0=tibble(var=NA,cv=NA,skew=NA,kurt=NA,shapiro=NA)
  tb1=tb0
  for(i in c){
    m=select(tb,i) |> drop_na() |> as.matrix()
    if(F){
      hist(m,main=i,
           xlab=str_c('cv:',format(sd(m)/mean(m)),
                      ' skew:',format(skew(m)),
                      ' kurt:',format(kurt(m))))
    }
    tb1[1,1]=i
    tb1[1,2]=format(sd(m)/mean(m))
    tb1[1,3]=format(skew(m))
    tb1[1,4]=format(kurt(m))
    tb1[1,5]=format(shapiro.test(m)$p.value)
    tb0=bind_rows(tb0,tb1)
  }
  tb0[-1,] |> kable('simple')
}

#see distribution of all numeric variables by all categories
distsByCat=function(tb){
  options(digits=2)
  par(mfrow=c(3,3))
  v=select(tb,is.numeric) |> colnames()
  c=select(tb,is.factor) |> colnames()
  tb0=tibble(n_var=NA,c_var=NA,cv=NA,skew=NA,kurt=NA,shapiro=NA)
  tb1=tb0
  for(i in v){
    for(j in c){
      f=tb[j] |>
        unlist() |>
        levels()
      for(k in f){
        m=tb  |> 
          filter(.data[[j]]==k) |>
          select(i) |>
          drop_na() |>
          as.matrix()
        if(F){
          hist(m,main=str_c(i,'\n',j,':',k),
               xlab=str_c('cv:',format(sd(m)/mean(m)),
                          ' skew:',format(skew(m)),
                          ' kurt:',format(kurt(m))))
        }
        tb1[1,1]=i
        tb1[1,2]=str_c(j,':',k)
        tb1[1,3]=format(sd(m)/mean(m))
        tb1[1,4]=format(skew(m))
        tb1[1,5]=format(kurt(m))
        tb1[1,6]=format(shapiro.test(m)$p.value)
        tb0=bind_rows(tb0,tb1)    
        
      }
    }
  }
  tb0[-1,] |> kable('simple') 
}

#see correlation easily
easyCR=function(tb){
  select(tb,is.numeric) |> 
    cor(use='complete.obs') |> 
    #print.table(digits=2) |>
    kable('simple')
}

#cramerV
crv=function(c1,c2){
  c1=as.matrix(c1)
  c2=as.matrix(c2)
  crs=table(c1,c2)
  chisq=chisq.test(crs)$statistic
  n=sum(crs)
  v=sqrt(chisq/(n*(min(nrow(crs),ncol(crs))-1)))
  names(v)='cramerV'
  return(v)
}

#cramerV for table or matrix
crvm=function(m){
  chi2=chisq.test(m)$statistic
  n=sum(m)
  v=sqrt(chi2/(n*(min(nrow(m),ncol(m))-1)))
  names(v)='cramerV'
  return(v)
}

#see Cramer's V in all couple of categorical valuables
crvall=function(tb){
  options(digits=2)
  par(mfrow=c(2,2))
  c=select(tb,is.factor) |> colnames()
  tb0=tibble(var1=c(NA),var2=c(NA),cramerV=c(NA))
  tb1=tb0
  for(i in c){
    for(j in c){
      if(i!=j){
        t=tibble(tb[i],tb[j]) |> table()
        cr=crv(tb[i],tb[j])
        if(F){
          mosaicplot(t,main='',
                     sub=str_c('cramer\'sV:',format(cr)))
        }
        tb1[1,1]=i
        tb1[1,2]=j
        tb1[1,3]=cr
        tb0=bind_rows(tb0,tb1)
      }
    }
  }
  tb0[-1,] |> kable('simple')
}

#Mantel-Haenszel
#(c1,c2,c3) independence of c1 and c2 on c3 fixed
crvmh=function(c1,c2,c3){
  c1=as.matrix(c1)
  c2=as.matrix(c2)
  c3=as.matrix(c3)
  crs=table(c1,c2,c3)
  mh=mantelhaen.test(crs)[1]$statistic
  n=sum(crs)
  v=sqrt(mh[1]/(n*(min(nrow(crs),ncol(crs))-1)))
  names(v)='cramerV_MH'
  return(v)
}

#summary by c1
agr1c=function(tb,v,c1){
  f=as.formula(str_c(v,'~',c1))
  a=aggregate(f,tb,summary)
  m=aggregate(f,tb,mean)
  s=aggregate(f,tb,sd)
  b=bind_cols(a[[1]],a[[2]],s[[2]],s[[2]]/m[[2]],)
  c=names(a)
  colnames(b)[1]=c[1]
  colnames(b)[8]='sd'
  colnames(b)[9]='cv'
  print(c[2])
  kable(b,'simple')
}

#summary by c1,c2
agr2c=function(tb,v,c1,c2){
  f=as.formula(str_c(v,'~',c1,'+',c2))
  a=aggregate(f,tb,summary)
  m=aggregate(f,tb,mean)
  s=aggregate(f,tb,sd)
  b=bind_cols(a[[1]],a[[2]],a[[3]],s[[3]],s[[3]]/m[[3]])
  c=names(a)
  colnames(b)[1]=c[1]
  colnames(b)[2]=c[2]
  colnames(b)[9]='sd'
  colnames(b)[10]='cv'
  print(c[3])
  kable(b,'simple')
}

#see eta in all numeric vars by all categorical vars
etaall=function(tb){
  options(digits=2)
  par(mfrow=c(2,2))
  v=select(tb,is.numeric) |> colnames()
  c=select(tb,is.factor) |> colnames()
  tb0=tibble(n_var=c(NA),c_var=c(NA),eta=c(NA))
  tb1=tb0
  for(i in v){
    for(j in c){
      f=as.formula(str_c(i,'~',j))
      anv=aov(f,tb)
      # strictly aov can be used in balanced design and 1wayANOVA
      anv=summary(anv) |> unlist()   
      if(F){
        boxplot(f,tb,main=j,ylab=i,
                xlab=str_c('eta:',
                           format(anv['Sum Sq1']/(anv['Sum Sq1']+anv['Sum Sq2']))))
      }
      
      #use car::Anova
      #anv=Anova(lm(f,tb),type=2)
      #boxplot(f,tb,main=j,ylab=i,
      #    xlab=str_c('eta:',
      #               format(anv$'Sum Sq1'[1]/
      #                        (anv$'Sum Sq1'[1]+anv$'Sum Sq2'[2]))))
      
      tb1[1,1]=i
      tb1[1,2]=j
      tb1[1,3]=format(anv['Sum Sq1']/(anv['Sum Sq1']+anv['Sum Sq2']))
      tb0=bind_rows(tb0,tb1)   
    }
  }
  tb0[-1,] |> kable('simple')
}
```


# pretreatment
see table data and change variable type appropriately, reject row has missing, reject outlier and make summary table
```{r}
tb=penguins
glimpse(tb)

tb[!complete.cases(tb),]

tb %<>% drop_na()

tb$year=as.factor(tb$year)
tb |>  
  select(is.factor) -> tbc
tb |> select(is.numeric) -> tbn

par(mar=c(3,8,3,3))
boxplot(scale(tb[colnames(tbn)]),horizontal=T,las=2)

tb=outliers(tb)

smArs(tb)
```


# discription of variable to do EDA

```
Verification of the coefficient of variation, variables with relatively low or high variance will be treated as constants based on their mean and excluded from the model variables.

Verification of normality for variables to be included in the model, estimate the process by which variable values are generated and confirm the applicability of tests that require normality.

Exclude variables that are nearly uncorrelated with all other variables.

Exclude variables that are fully explained by a combination of other variables.

Convert categorical variables into binary dummy variables by level (for functions that only accept numeric arguments).
```

## numeric variables
```{r}
dists(tb)

summary(tb$body_mass_g)

par(mfrow=c(1,2))
hist(tb$body_mass_g,main='Histgram',xlab='body_mass_g')
density(tb$body_mass_g,na.rm=T) |> plot(main='Distribution',xlab='body_mass_g')

grid.arrange(
  qplot(body_mass_g,data=tb,main='Histgram',binwidth=500,xlab='body_mass_g'),
  qplot(body_mass_g,data=tb,geom='density',main='Distribution',xlab='body_mass_g'),
  ncol=2
)
```


## see distribution family
### continuous distribution
```{r}
#see continuous distribution family by ks-test
distCnt=function(x){
  l=min(x)
  u=max(x)
  m=mean(x)
  s2=var(x)
  cat('mean:',mean(x),
      'sd:',sd(x),
      'skew:',skew(x),
      'kurt:',kurt(x),
      '\n')
  cat('ks-test p-value\n')
  cat('uniform : skew(0),kurt(1.8),[l:',l,',u:',u,'] p:',
      ks.test(x,'punif',l,u)$p.value,'\n')
  cat('normal : skew(0),kurt(3),[m:',m,',s2:',s2,'] p:',
      ks.test(x,'pnorm',m,s2)$p.value,'\n')
  if(l>0){
    cat('log normal :[m:',mean(log(x)),',s2:',var(log(x)),'] p:'
        ,ks.test(x,'plnorm',mean(log(x)),var(log(x)))$p.value,'\n')
  }
  if(l>=0){
    cat('exponential : skew(2),kurt(9),[p:,',1/m,'] p:',
        ks.test(x,'pexp',1/m)$p.value,'\n') #m=1/l
    cat('gamma : skew(',2/m^.5,'),kurt(',6/m+3,'),[a:,',m^2/s2,'p:',m/s2,'] p:',
        ks.test(x,'pgamma',m^2/s2,m/s2)$p.value,'\n') #m=a/l,s2=a/l^2
  }
  if(l>=0 && u<=1){
    a=(m^2-m^3)/s2-m; b=a*(1-m)/m #m=a/(a+b),s2=ab/((a+b)^2*(a+b+1))
    cat('beta :[a:',a,'b:',b,'] p:',
        ks.test(x,'pbeta',a,b)$p.value,'\n')
  }
}
```

```{r,class.source="fold-show"}
#uniform dist. x(-Inf,Inf)
x=runif(20,-2,3)
distCnt(x)
#normal dist. x(=inf,Inf)
x=rnorm(20,3,2)
distCnt(x)
#log normal dist. x(0,Inf)
x=rlnorm(20,2,1)
distCnt(x)
#exponentiol dist. x[0,Inf)
x=rexp(20,0.2)
distCnt(x)
#gamma dist. x[0,Inf)
x=rgamma(20,3,rate=0.2)
distCnt(x)
#beta dist. x[0,1]
x=rbeta(20,2,3)
distCnt(x)
```


### discrete distribution
```{r}
#see discrete distribution family by chisq-test
distDsc=function(x){
  x=outlier(x)
  l=min(x)
  if(l<0) return('x has value less than 0')
  u=max(x)
  m=mean(x)
  s2=var(x)
  cat('mean:',mean(x),
      'sd:',sd(x),
      'skew:',skew(x),
      'kurt:',kurt(x),
      '\n')
  f=table(factor(x,levels=0:u))

  cat('chisq-test p-value\n')

  #binomial
  p0=1-s2/m; n=round(m^2/(m-s2)) #m=np, s2=np(1-p)
  if(p0>=0 && n>0){
    for(i in 0:n){
      xx=paste(i)
      if(is.na(f[xx])) f[xx]=0
    }
    x1=names(f) |> as.numeric()
    p=dbinom(x1,n,p0)
    pv=suppressWarnings(chisq.test(x=f,p=p)$p.value)
    cat('binomial : [n:',n,',p:',p0,'] p:',pv,'\n')
  }
  
  f=table(factor(x,levels=0:u))
  x1=names(f) |> as.numeric()
  f[u+2]=0
  
  #poisson
  l=m #m=l, s2=l^2 
  p=dpois(x1,l)
  p[u+2]=1-sum(p)
  pv=suppressWarnings(chisq.test(x=f,p=p)$p.value)
  cat('poisson : skew(',1/m^.5,'),kurt(',1/m+3,'),[l:',l,'] p:',pv,'\n')  

  #geometry
  p0=1/(m+1) #m=(1-p)/p, s2=(1-p)/p^2 
  p=dgeom(x1,p0) 
  p[u+2]=1-sum(p)
  pv=suppressWarnings(chisq.test(x=f,p=p)$p.value)
  cat('geometry : skew(',(2-p0)/(1-p0)^.5,'),kurt(',6+p0^2/(1-p0),'),[p:',p0,'] p:',pv,'\n')

  #negeative binomial
  p0=m/s2 ;r=m^2/(s2-m) #m=r(1-p)/p, s2=r(1-p)/p^2
  if(p0>=0 && r>0){
    p=dnbinom(x1,r,p0) 
    p[u+2]=1-sum(p)
    pv=suppressWarnings(chisq.test(x=f,p=p)$p.value)
    cat('n-binomial : [r:',r,'],[p:',p0,'] p:',pv,'\n')
  }
}
```

```{r,class.source="fold-show"}
#binomial dist. x[0,n]
#p: event probability, n: #trial, x: #event occurence
x=rbinom(20,10,0.3)
distDsc(x)
#poisson dist. x[0,Inf)
#l: mean event occurence, x: #event occurence
x=rpois(20,3)
distDsc(x)
#geometry dist. x[0,Inf)
#p: event probability, x: #failure to occur first event
x=rgeom(20,0.3)
distDsc(x)
#negeative binomial dist. x[r,Inf)
#p: event probability, r: #event, x: #failure to occur r times events
x=rnbinom(20,3,0.5)
distDsc(x)
```


## make normal distribution
### skewed to the right
```{r}
normality=function(x){
  #density(x) |> plot(main='')
  cat('skewness: ',skew(x),', kurtosis: ',kurt(x),', Shapiro-Wilk test: ',shapiro.test(x)$p.value,'\n')
}

cat('x0\n')
x0=rnorm(300,10,1)
normality(x0)

cat('x0^2\n')
xr2=x0^2
normality(xr2)

cat('x0^3\n')
xr3=x0^3
normality(xr3)

cat('log x0^2\n')
normality(log(xr2))

cat('log x0^3\n')
normality(log(xr3))

cat('x=x0^2\n')
for(i in c(1,1.5,2,2.5,3,3.5,4)){
  cat('x^ 1/',i,'\n')
  normality(xr2^(1/i))
}

cat('x=x0^3\n')
for(i in c(1,1.5,2,2.5,3,3.5,4)){
  cat('x^ 1/',i,'\n')
  normality(xr3^(1/i))
}
```

### skewed to the left
```{r}
cat('x0^ 1/2\n')
xl2=x0^(1/2)
normality(xl2)

cat('x0^ 1/3\n')
xl3=x0^(1/3)
normality(xl3)

cat('exp x0^ 1/2\n')
normality(exp(xl2))

cat('exp x0^ 1/3\n')
normality(exp(xl3))

cat('x=x0^ 1/2\n')
for(i in c(1,1.5,2,2.5,3,3.5,4)){
  cat('x^ ',i,'\n')
  normality(xl2^i)
}

cat('x=x0^ 1/3\n')
for(i in c(1,1.5,2,2.5,3,3.5,4)){
  cat('x^ ',i,'\n')
  normality(xl3^i)
}
```



## categorical variable
```{r}
#change type to ordered factor
head(tb$year)
tb$year=ordered(tb$year,levels=c('2007','2008','2009'))
head(tb$year)
tb$year |> as.numeric()

#frequency
xtabs(~species,tb)
#table(tb$species)
xtabs(~species,tb) |> prop.table()

xtabs(~species,tb) |> barplot(xlab='species')
par(mar=c(3,5,3,3)) #dowm,left,up,right
xtabs(~species,tb) |> barplot(xlab='species',horiz=T,las=2)

qplot(tb$species,xlab='species')
qplot(tb$species,xlab='species')+coord_flip()
```


```{r}
xtabs(~species+sex,tb) #(row+col)
#table(tb$species,tb$sex) #(row,col)

xtabs(~species+sex+year,tb)
ftable(species~sex+year,tb) #(cols~rows) or row is more

xtabs(~species,tb) %>% as_tibble()
xtabs(~species+sex,tb) %>% as_tibble()
xtabs(~species+sex+year,tb) %>% as_tibble()

xtabs(~species+sex,tb) |> addmargins()
xtabs(~species+sex,tb) |> margin.table(1) #sum by row
xtabs(~species+sex,tb) |> margin.table(2) #sum by col
xtabs(~species+sex,tb) |> margin.table() #sum all exclude NA

xtabs(~species+sex,tb) |> prop.table(1) #proportion by row
xtabs(~species+sex,tb) |> prop.table(2) #proportion by col
xtabs(~species+sex,tb) |> prop.table() #proportion in all

par(mfrow=c(2,1))
xtabs(~species+sex,tb) |> barplot(horiz=T,las=2)
xtabs(~sex+species,tb) |> barplot(horiz=T,las=2)

xtabs(~species+sex,tb) |> barplot(beside=T,horiz=T,las=2)
xtabs(~sex+species,tb) |> barplot(beside=T,horiz=T,las=2)

xtabs(~species+sex,tb) |> prop.table(2) |> barplot(horiz=T,las=2)
xtabs(~sex+species,tb) |> prop.table(2) |> barplot(horiz=T,las=2)

par(mfrow=c(1,1))
xtabs(~species+sex,tb) |> plot(main='')
ftable(species~sex+year,tb) |> as.table() |> plot(main='')
```



## aggregation
```{r}
aggregate(body_mass_g~species,tb,mean)
aggregate(body_mass_g~species+sex,tb,mean)

agr1c(tb,'body_mass_g','species')
agr2c(tb,'body_mass_g','species','sex')
```


```{r}
distsByCat(tb)
```

```{r}
boxplot(body_mass_g~sex,tb)
boxplot(body_mass_g~species,tb)
```

```{r}
qplot(body_mass_g,data=tb,facets=~sex,main='Histgram',binwidth=500,xlab='body_mass_g')
qplot(body_mass_g,data=tb,geom='density',facets=~sex,main='Distribution',xlab='body_mass_g')
```

```{r}
qplot('',body_mass_g,data=tb,geom='boxplot',facets=~sex,xlab='')
qplot('',body_mass_g,data=tb,geom='violin',facets=~sex,xlab='')
```

```{r}
qplot(body_mass_g,data=tb,facets=~sex+species,main='Histgram',binwidth=200,xlab='body_mass_g')
qplot(body_mass_g,data=tb,geom='density',facets=~sex+species,main='Distribution',xlab='body_mass_g')
```

```{r}
par(mfrow=c(1,1))
par(mar=c(3,8,3,3)) #dowm,left,up,right
boxplot(body_mass_g~species+sex,tb,horizontal=T,las=2,ylab='')
```

```{r}
par(mar=c(3,10,3,3)) #dowm,left,up,right
boxplot(body_mass_g~year+species+sex,tb,horizontal=T,las=2,ylab='')
```

```{r}
qplot('',body_mass_g,data=tb,geom='boxplot',facets=~sex+species,xlab='')
qplot('',body_mass_g,data=tb,geom='violin',facets=~sex+species,xlab='')
```

```{r}
qplot(species,body_mass_g,data=tb,facets=~sex,geom='boxplot')
qplot(species,body_mass_g,data=tb,facets=~sex,geom='violin')
```

```{r}
qplot(year,body_mass_g,data=tb,facets=~species+sex,geom='boxplot')
qplot(year,body_mass_g,data=tb,facets=~species+sex,geom='violin')
```


## test group difference
```
In EDA, use effect size rather than p-values for testing.

Test Statistic = Function of Effect Size × Function of Sample Size (Reliability of Effect Size)

Effect Size:
    d Family: Standardized mean difference between groups for one variable, such as Cliff’s delta or Cohen’s d.
    r Family: Measure of the strength of association between two variables, such as Cramer’s V, Eta-squared (η²), or correlation coefficient.

p-Value: The probability that the test statistic exceeds a given significance level.

Traditional test statistics depend on sample size:
    With small sample sizes, even if a calculated effect size is large, the test statistic may not be significant, making it difficult to detect significant differences.
    With large sample sizes, even when the effect size is close to zero, the test statistic can be large, making it easier to detect significant differences.

The appropriate sample size should be calculated based on meaningful differences, significance level, and the power of the analysis method.

To determine whether a difference or association exists using p-values derived from test statistics, the data must be obtained with an appropriate sample size. However, when analyzing existing data, p-value testing is unnecessary. Instead, examine whether the effect size and the confidence interval of the difference or correlation exclude zero. Use d-family effect size as a substitute for the t-test and r-family effect size as a substitute for the chi-square test and F-test in ANOVA.

Guidelines for interpreting effect size can be found here: https://kamesan-kamesan.com/effect-size/
```
```
Glass's Δ
size of the difference from population, given population mean x0, sd s0
```
$$Δ=(\bar{x}-x_0)/s_0$$
```
Cohen's d
size of the difference between non coresponding group A and B
```
$$d=(\bar{x_a}-\bar{x_b})/s$$
$$s=\sqrt{((n_a-1)s_a^2+(n_b-1)s_b^2)/(n_a+n_b-2)}$$
```
Hedge's g
correction of Cohen's d for small sample
```
$$g=d*(1-3/(4(n_a+n_b)-9))$$
```
diff's δ
size of the difference between coresponding group A and B
```
$$δ=(\#(x_a>x_b)-\#(x_a<x_b))/n_a/n_b$$

```{r}
# 2 groups of numeric variable 
boxplot(body_mass_g~sex,tb)

t.test(body_mass_g~sex,tb)
wilcox.test(body_mass_g~sex,tb)

m0=mean(tb$body_mass_g)
s0=sd(tb$body_mass_g)
cat("Glass's Δ ",
    ((filter(tb,sex=='male') |> select(body_mass_g) |> unlist() |> mean())-m0)/s0)
cliff.delta(body_mass_g~sex,tb)
cohen.d(body_mass_g~sex,tb)
cohen.d(body_mass_g~sex,tb,hedges.correction=T)

anv=aov(body_mass_g~sex,tb) |>
  summary() |> unlist()
cat('eta: ',anv['Sum Sq1']/(anv['Sum Sq1']+anv['Sum Sq2']),'\n')
```

\

```{r}
# more than 2 groups of numeric variable 
boxplot(body_mass_g~species,tb)
anv=aov(body_mass_g~species,tb) |> print() |> 
  summary() |> unlist()
cat('eta: ',anv['Sum Sq1']/(anv['Sum Sq1']+anv['Sum Sq2']),'\n')
```

\

```{r}
# 2 groups of categorical variable, difference of rario
select(tb,species,year) |>
  mutate(across(everything(),as.character)) |> # factor -> character
  filter(year %in% c(2007,2009),
         species %in% c('Chinstrap','Gentoo')) |> 
  table() |> print() |> 
  prop.table(1) -> pt # argument 1 for proportion by row

pt
mosaicplot(pt,main='')
prop.test(pt)
```

\

## correlation, association
```{r}
cor(tbn,use='complete.obs')

easyCR(tbn)

plot(tbn)
```

```{r}
corrplot(cor(tbn,use='complete'),
         diag=F,type='upper',method='ellipse',order='AOE',cl.pos='n')
```

```{r}
etaall(tb)

tbnm=tbn[tb$sex=='male',]
cor(tbnm,use='complete.obs')
plot(tbnm)
```

```{r}
corrplot(cor(tbnm,use='complete'),
         diag=F,type='upper',method='ellipse',order='AOE',cl.pos='n')
```

```{r}
tbnf=tbn[tb$sex=='female',]
cor(tbnf,use='complete.obs')
plot(tbnf)
```

```{r}
corrplot(cor(tbnf,use='complete'),
         diag=F,type='upper',method='ellipse',order='AOE',cl.pos='n')
```


```{r}
# groups of categorical variable, see independence
mosaicplot(year~species,tb,main='')
select(tb,species,year) |>
  table() |> print() |> 
  crvm()

#independence test
xtabs(~species+sex,tb) |> chisq.test()
xtabs(~species+sex,tb) |> fisher.test()

xtabs(~species+sex,tb) |> crvm()
crvall(tb)

#Mantel-Haenszel test
xtabs(~species+sex+year,tb) |> mantelhaen.test()
#(c1,c2,c3) independence of c1 and c2 on c3 fixed
```


```
In relationships among 3 variables, it is important to examine interactions, especially negative interactions.
```

```{r}
# 3 categorical variables
mosaicplot(year~species+sex,tb,main='')

# categorical vs categorical vs numeric
# see eta from ANOVA, see interaction term
boxplot(body_mass_g~sex+species,tb)
anv=aov(body_mass_g~sex+species+sex*species,tb) # or body_mass_g~sex*species
summary(anv)

par(mfrow=c(1,2))
# categorical vs numeric vs numeric
# see slope of stratified scatter, see eta from ANCOVA, see interaction term
plot(body_mass_g~bill_length_mm,tb,pch=as.integer(species))
plot(body_mass_g~bill_length_mm,tb,col=as.integer(species))
acv=aov(body_mass_g~bill_length_mm*species,tb) # ANCOVA
summary(acv)

plot(select(tb,is.numeric),col=as.integer(tb$species))

# 3 numeric variables
# correlation
cor(select(tb,body_mass_g,bill_length_mm,flipper_length_mm))

# partial correlation
pcor(select(tb,body_mass_g,bill_length_mm,flipper_length_mm))$estimate
```


### log linear model
```{r}
#log linear model
tbl=xtabs(~species+sex+year,tb)
 #frequency f(i*j*k) on table (a*b*c)
mdl1=loglm(~species,tbl) #log f=m+a
mdl2=loglm(~species+sex,tbl) #log f=m+a+b
mdl3=loglm(~species+sex+year,tbl) #log f=m+a+b+c
mdl4=loglm(~species*sex,tbl) #log #log f=m+a+b+ab
mdl5=loglm(~species*sex+year,tbl) #log f=m+a+b+ab+c
mdl6=loglm(~(species+sex)*year,tbl) #log f=m+a+b+c+ac+bc
mdl7=loglm(~species*sex+species*year+sex*year,tbl) #log f=m+a+b+c+ab+ac+bc
mdl8=loglm(~species*sex*year,tbl) #log f=m+a+b+c+ab+ac+bc+abc

anova(mdl1,mdl2,mdl3,mdl4,mdl5,mdl6,mdl7,mdl8,test='Chisq')

frq=xtabs(~species+sex+year,tb) |> as_tibble()
mdl0=glm(n~species+sex+year+
           species:sex+species:year+sex:year+
           species:sex:year,frq,family=poisson('log'))
mdl0
bstmdl=step(mdl0)
bstmdl

anova(bstmdl,mdl0,test='Chisq')
```

