---
title: "Medical data analysis"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show 
#date: "`r Sys.Date()`"

---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=T,warning=F,message=F)

suppressWarnings(
  suppressMessages(
    suppressPackageStartupMessages({
      library(stats)
      library(MASS)
      library(tidyverse)
      library(magrittr)
      library(gridExtra)
      
      library(palmerpenguins)　#サンプルデータ用

    })
  )
)
options(scipen=100,digits=3)
```

## see each variable normality
decide whether to choose parametric test or rank based (nonparamatric) test
```{r}
tb=penguins |> drop_na()
glimpse(tb)
tb$year=as.factor(tb$year)

density(tb$body_mass_g) |> plot()
qqnorm(tb$body_mass_g);qqline(tb$body_mass_g)
shapiro.test(tb$body_mass_g)
ks.test(tb$body_mass_g,'pnorm')
```


## difference of 2 group's mean
```{r}
aggregate(body_mass_g~sex,tb,'mean')
qplot(sex,body_mass_g,data=tb,geom='violin')
```

### t-test
```{r}
t.test(body_mass_g~sex,tb)
```

### Wilcoxon rank test (Mann–Whitney U test)
```{r}
wilcox.test(body_mass_g~sex,tb,exact=F)
```


## any difference of more than 2 group's mean
```{r}
aggregate(body_mass_g~species,tb,'mean')
qplot(species,body_mass_g,data=tb,geom='violin')
```

#### see variance equality in normal distribution
```{r}
bartlett.test(body_mass_g~species,tb)
```

### ANOVA
```{r}
aov(body_mass_g~species,tb) |> summary()
```

### multiple comparison of more than 2 group's mean after ANOVA
```{r}
pairwise.t.test(tb$body_mass_g,tb$species)
```


#### see variance equality in unnormal distribution 
```{r}
fligner.test(body_mass_g~species,tb)
```

### Kruskal test
```{r}
kruskal.test(body_mass_g~species,tb)
```

### multiple comparison of more than 2 group's mean after Kruskal test
```{r}
pairwise.wilcox.test(tb$body_mass_g,tb$species)
```



## difference of each subject's mean at 2 points
```{r}
tb0=sleep
glimpse(tb0)
names(tb0)[2]='at'
tb1=subset(tb0,at==1)
tb2=subset(tb0,at==2)
tb=tb0
levels(tb$at)=c('before','after')

aggregate(extra~at,tb,'mean')
qplot(at,extra,data=tb,geom='violin')
```

### paired t-test
```{r}
t.test(tb1$extra,tb2$extra,paired=T)
```

### Wilcoxon signed-rank test
```{r}
wilcox.test(tb1$extra,tb2$extra,paired=T,exact=F)
```


## difference of each subject's mean at more than 2 points
```{r}
tb3=tb1
tb3$at='3'
tb3$extra=tb1$extra+tb2$extra+rnorm(nrow(tb3),0,1)
tb=rbind(tb1,tb2,tb3)
levels(tb$at)=c('t1','t2','t3')

aggregate(extra~at,tb,'mean')
qplot(at,extra,data=tb,geom='violin')
```

### repeated ANOVA
```{r}
aov(extra~at+Error(ID/at),tb) |> summary()
```

### multiple comparison of more than 2 group's mean after ANOVA
```{r}
pairwise.t.test(tb$extra,tb$at,paired=T)
```


### Freidman test
```{r}
friedman.test(extra~at|ID,tb)
```

### multiple comparison of more than 2 group's mean after Friedman test
```{r}
pairwise.wilcox.test(tb$extra,tb$at,paired=T,exact=F)
```



## cross table analysis
### ratio difference of 2 groups, chi square test, Fisher exact test
```{r}
mx=xtabs(~ap+y,bacteria) |> print()
prop.test(mx)
chisq.test(mx)
fisher.test(mx) # for small sample
```


### any ratio difference of more than 2 groups, chi square test
```{r}
tb=penguins |> drop_na()
xtabs(~species+sex,tb) |> print() |> chisq.test()
```

#### multiple comparison of more than 2 group's ratio
```{r}
xtabs(~species+sex,tb) |> print() |> pairwise.prop.test()
```


### inverse weighted cross table, unbalance #items by group to be balanced
```{r}
tb=bacteria
glimpse(tb)
library(survey)

xtabs(~ap+y,tb) |> print() |>
  chisq.test() |> print()
cat('\n')

nk=table(tb$hilo) # Frequency of each level in the confounding factor
tb$wt=1/nk[tb$hilo] # Assign inverse frequency as weight

dsn=svydesign(ids=~1,data=tb,weights=~wt)
iwtbl=svytable(~ap+y,design=dsn) |> addmargins() |> print()
iwtbl*nrow(tb)/2
svychisq(~ap+y,design=dsn)
```


### multi cross table, ratio difference of groups in each category table, Mantel Haenszel test
```{r}
xtabs(~ap+y,tb) |> print() |> chisq.test()
cat('\n hi\n')
xtabs(~ap+y,filter(tb,hilo=='hi')) |> print() |> chisq.test()
cat('\n lo\n')
xtabs(~ap+y,filter(tb,hilo=='lo')) |> print() |> chisq.test()

mantelhaen.test(tb$ap,tb$y,tb$hilo)
```


### ratio difference of each subject's at 2 points, McNemar test
```{r}
tb0=sleep
glimpse(tb0)
names(tb0)[2]='at'
xtabs(~at+(extra>0),tb0) |> print() |> mcnemar.test()
```


### ratio difference of each subject's at more than 2 points, Cochran Q test
```{r}
tb1=subset(tb0,at==1)
tb2=subset(tb0,at==2)
tb3=tb1
tb3$at='3'
tb3$extra=tb1$extra+tb2$extra+rnorm(nrow(tb3),0,1)
tb=rbind(tb1,tb2,tb3)
levels(tb$at)=c('t1','t2','t3')
glimpse(tb)

library(DescTools)
xtabs(~at+(extra>0),tb) |> print() |> CochranQTest() 
```


### ratio difference with intervention ordered by strength, Cochran-Armitage trend test
```{r}
names(tb)[2]='order'
levels(tb$order)=c('weak','mid','strong')
glimpse(tb)

library(DescTools)
xtabs(~order+(extra>0),tb) |> print() |> CochranArmitageTest() 
```



### ROC
```{r}
library(pROC)
glimpse(aSAH)
aggregate(s100b~outcome,aSAH,'mean')
qplot(outcome,s100b,data=aSAH,geom='violin')

roc1=roc(outcome~age,aSAH,ci=T) |> print()
roc2=roc(outcome~s100b,aSAH,ci=T) |> print()

par(mfrow=c(1,2))
plot(roc1)
plot(roc2)

grid.arrange(ggroc(roc1,legacy.axes=T)+geom_abline(col='gray'),
             ggroc(roc2,legacy.axes=T)+geom_abline(col='gray'),
             nrow=1)

roc.test(roc1,roc2)
```


## variables relation
```{r}
tb0=penguins |> drop_na()
tb0$year=as.factor(tb0$year)
tb=subset(tb0,species=='Adelie' & sex=='male')
tbn=select(tb,is.numeric)
```

### correlation
```{r}
cor(tbn)
cor(tbn,method='kendall')

library(psych)
corr.test(tbn)
corr.test(tbn,method='kendall')
```

### partial correlation
```{r}
library(ppcor)
pcr=pcor(tbn)
pcr$estimate
pcr$p.value

pcr=pcor(tbn,method='kendall')
pcr$estimate
pcr$p.value
```


## effect of variables, evaluate model
```{r}
fit=lm(body_mass_g~flipper_length_mm,tb0)
anova(fit)
coef(fit)
confint(fit)
```

```{r}
fit=lm(body_mass_g~flipper_length_mm+species,tb0)
anova(fit)
coef(fit)
confint(fit)
```

```{r}
fit=lm(body_mass_g~species+sex,tb0)
anova(fit)
coef(fit)
confint(fit)
```

```{r}
fit=lm(body_mass_g~.,tb0)
anova(fit)
coef(fit)
confint(fit)
```

### see multicolinearlity by VIF
explanation quantity of a exp.var is explained by other exp.vars  
```{r}
fit=lm(flipper_length_mm~.-body_mass_g,tb0) # all variable except objective variable
vif=1/(1-summary(fit)$r.squared)
vif # >10 reduce the variable
```


## model copmparison
### liklihood ratio test
```{r}
fit1=lm(body_mass_g~flipper_length_mm,tb0) # simple model
fit2=lm(body_mass_g~flipper_length_mm+bill_depth_mm,tb0) # more general model

l1=logLik(fit1) # model log liklihood
l2=logLik(fit2)
ld=-2*(l1-l2) # models log liklihood ratio, follow chi2(df=1)
1-pchisq(ld,1) # <0.05 fit2 is better, >0.05 fit1(simple model) is better
```

### AIC
```{r}
fit=lm(body_mass_g~.,tb0)
#AIC=-2*log(model likihood)+2*#parameter
AIC(fit)
step(fit)
```


## prediction
```{r}
fit=lm(body_mass_g~flipper_length_mm,tb0)
l=min(tb0$flipper_length_mm)
u=max(tb0$flipper_length_mm)
conf=predict(fit,list(flipper_length_mm=l:u),interval='confidence')
pred=predict(fit,list(flipper_length_mm=l:u),interval='prediction')
```


## logistic regression
```{r}
fit=glm(sex~body_mass_g+bill_depth_mm,tb0,family=binomial('logit'))
anova(fit)
coef(fit)
confint(fit)
cat('\n odds ratio\n')
exp(coef(fit)[-1]) # odds ratio
exp(confint(fit)[-1,])
```

```{r}
fit=glm(sex~.,tb0,family=binomial('logit'))
anova(fit)
coef(fit)
confint(fit)
cat('\n odds ratio\n')
exp(coef(fit)[-1]) # odds ratio
exp(confint(fit)[-1,])

cat('\n reduce variable by AIC\n')
step(fit)
```



## IPTW, propensity score matching by IPW
```{r}
library(PSweight)
tb=VA
glimpse(tb)
tb$status=as.factor(tb$status)
qplot(treat,stime,data=tb,geom='violin')

f=treat~age+Karn+diag.time+cell+prior
ps=SumStat(ps.formula=f,data=tb,weight='IPW')

plot(ps,type='density')
plot(ps,type='balance')

iptw=PSweight(ps.formula=f,yname='stime',data=tb,weight='IPW')
summary(iptw,type='DIF')
```



## survival time analysis
Analysis for data has end or censoring of observation 

Kaplan-Meyor estimate
$KM=\prod _i\left(1-e_i/n_i\right)$  
$t_i=0,...T$ time has interval  
$n_i$ number at risk at time $t_i$  
$e_i$ number of events at time $t_i$  
$t_i$ at KM=0.5 is time number of no-event is 50% of initial number at risk  

```{r}
library(survival)
tb=lung
glimpse(tb)
srv=Surv(tb$time,tb$status==2) # time:observation time, status 1:censored, 2:event

km=survfit(srv~1,tb) |> print()
plot(km,xlab='time',ylab='survival probability')

km=survfit(srv~sex,tb) |> print()
plot(km,xlab='time',ylab='survival probability',col=c('blue','red'))
legend('topright',legend = c("Male", "Female"),col = c('blue','red'),lwd=2)
```

### survivality difference without weight, log rank test
```{r}
survdiff(srv~sex,tb)
```

### survivality difference with # at risk weight, generalized Wilcoxon test
```{r}
survdiff(srv~sex,tb,rho=1)
```

### KM for data has start time and end time
```{r}
tb=heart
glimpse(tb)
srv=Surv(tb$start,tb$stop,tb$event==1) # time1/2:start/end, event 0:censored, 1:event

km=survfit(srv~1,tb) |> print()
plot(km,xlab='time',ylab='survival probability')

km=survfit(srv~transplant,tb) |> print()
plot(km,xlab='time',ylab='survival probability',col=c('blue','red'))
legend('topright',legend = c("not transplant ", "transplant"),col = c('blue','red'),lwd=2)
```


## cox proportional hazard model

$h\left(x_i,t\right)=h_0\left(t\right)\exp\sum^p_j b_j x_{ij}$  
$x_i$ covariate vector of subject i  
$h_0(t)$ baseline hazard function  

```{r}
tb=lung
glimpse(tb)
srv=Surv(tb$time,tb$status==2)
```

### see propotional hazard by group
```{r}
km=survfit(srv~sex,tb) |> print()
plot(km,fun='cloglog',xlab='time',ylab='survival probability',col=c('blue','red'))
legend('topright',legend = c("Male", "Female"),col = c('blue','red'),lwd=2) #log-log plot

coxph(srv~sex,tb)
```

```{r}
coxph(srv~age,tb)
coxph(srv~sex+age,tb)
```

### model has other h0(t) by group  
```{r}
coxph(srv~age+strata(sex),tb)
```
