---
title: "Pretreatment table data"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show 
#date: "`r Sys.Date()`"

---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=T,warning=F,message=F,comment='')


suppressWarnings(
  suppressMessages(
    suppressPackageStartupMessages({
      library(stats)
      library(MASS)
      library(tidyverse)
      library(magrittr)
      library(RSQLite)
      library(nanoparquet)
      library(httr)
      library(httr2)
      library(jsonlite)
      library(palmerpenguins)　#サンプルデータ用
    })
  )
)
options(scipen=100,digits=3)
```

## input data from csv
```{r}
tb0=read_csv('penguins.csv')
slice_sample(tb0,n=10)

read_csv('penguins.csv',col_types=cols(.default='c')) |>  
  type_convert()

read_csv('penguinsSJ.csv',col_names=T,locale=locale(encoding='shift-jis'))
```

## output data as csv
```{r}
tb=tb0
write_csv(tb,'penguins.csv')

write.csv(tb,'penguinsSJ.csv',fileEncoding='shift-jis',row.names=F)
```

## see table
```{r}
glimpse(tb0)
nrow(tb0)
ncol(tb0)

tb0$year=as.factor(tb0$year)
```

## treat NA
```{r}
tb0[!complete.cases(tb0),]
filter(tb0,is.na(sex))

replace_na(tb0,list(sex='no'))
drop_na(tb0)
```

## see variable statistic
```{r}
min(tb0$bill_length_mm,na.rm=T)
max(tb0$bill_length_mm,na.rm=T)
range(tb0$bill_length_mm,na.rm=T)
sum(tb0$bill_length_mm,na.rm=T)
mean(tb0$bill_length_mm,na.rm=T)
median(tb0$bill_length_mm,na.rm=T)
fivenum(tb0$bill_length_mm,na.rm=T)

var(tb0$bill_length_mm,na.rm=T)
sd(tb0$bill_length_mm,na.rm=T)

unique(tb0$species)
distinct(tb0,species,island)
xtabs(~species,tb0)
xtabs(~species+island,tb0)
```


## convert numeric variables
```{r}
tb=drop_na(tb0)
select(tb,is.numeric) |> floor()
select(tb,is.numeric) |> ceiling()
select(tb,is.numeric) |> round(1)
select(tb,is.numeric) |> log()
select(tb,is.numeric) |> log10()
select(tb,is.numeric) |> scale() |> head(10)

select(tb,is.numeric) |> 
  mutate(across(everything(), ~(.-mean(.))/sd(.)))

select(tb,is.numeric) |> 
  mutate(across(everything(), ~(.-min(.))/(max(.)-min(.))))
```

## treat date,datetime
```{r}
today()
now()

tb=tribble(~date,~datetime,
            '20200123','20200123 112233',
            '20210124','20210124 122334',
            '20220125','20220125 132435')

mutate(tb,ymd=ymd(date),ymd_hms=ymd_hms(datetime))

mutate(tb,y=year(ymd(date)),m=month(ymd(date)),d=day(ymd(date)),
       H=hour(ymd_hms(datetime)),M=minute(ymd_hms(datetime)),S=second(ymd_hms(datetime)))
```


## prepare table, choose column
```{r}
select(tb0,species,bill_length_mm)
select(tb0,is.numeric)
select(tb0,!is.numeric)
```

## prepare table, choose row
```{r}
filter(tb0,species=='Adelie')
filter(tb0,species=='Adelie',sex=='male')

filter(tb0,bill_length_mm<35)
filter(tb0,bill_length_mm>55)
filter(tb0,between(bill_length_mm,40,50))

filter(tb0,str_detect(species,'del'))
filter(tb0,str_detect(species,'^Ch'))
filter(tb0,str_detect(species,'oo$'))
filter(tb0,str_detect(species,'[A-C]'))
filter(tb0,str_detect(species,'[A,C,G]'))
filter(tb0,str_detect(species,'o{2}'))
```

## prepare table, sort row
```{r}
arrange(tb0,bill_length_mm)
arrange(tb0,desc(bill_length_mm))
arrange(tb0,bill_length_mm,bill_depth_mm)
```

## prepare table, change values
```{r}
mutate(tb0,year=as.character(year))
mutate(tb0,year=as.character(year) |> as.numeric())

mutate_all(tb0,as.character)

mutate(tb0,id=row_number())
mutate(tb0,rank=min_rank(bill_length_mm))
mutate(tb0,rank=dense_rank(bill_length_mm))

mutate(tb0,sex=if_else(sex=='male','M','F'))
mutate(tb0,species=case_when(species=='Adelie'~'A',
                            species=='Chinstrap'~'C',
                            T~'G'))
```


## make table longer, make table wider
```{r}
tb=mutate(tb0,id=row_number()) |> 
  select(id,everything()) |> print()

select(tb,id,is.numeric) |> 
  pivot_longer(cols=c(bill_length_mm,
                      bill_depth_mm,
                      flipper_length_mm,
                      body_mass_g),
               names_to='arg',values_to='value')

tb1=select(tb,id,is.numeric)
pivot_longer(tb1,cols=colnames(tb1[-1]),names_to='arg',values_to='value')

tb1=select(tb,id,!is.numeric)
pivot_longer(tb1,cols=colnames(tb1[-1]),names_to='arg',values_to='value')

tb1=mutate_all(tb,as.character)
tb=pivot_longer(tb1,cols=colnames(tb1[-1]),names_to='arg',values_to='value') |> print()


pivot_wider(tb,names_from=arg,values_from=value)
```


## combine columns, split columns
```{r}
tb=unite(tb0,col=x,species,island,sep='@') |> print()

separate(tb,col=x,c('species','island'),sep='@')
```

## summary table
```{r}
tb=drop_na(tb0)
rowwise(tb) |> mutate(x=mean(c(bill_length_mm,bill_depth_mm)))

group_by(tb,species,sex) |>
  summarise(n=n())
group_by(tb,species,sex) |>
  summarise(n=n(),
            sum=sum(body_mass_g),
            mean=mean(body_mass_g)) |> 
  ungroup()
```


## string maipulation
```{r}
tb=select(tb0,!is.numeric)
mutate(tb,len=str_length(tb$species))
mutate(tb,x=str_to_lower(species))
mutate(tb,x=str_to_upper(species))
mutate(tb,x=str_c(species,'@',island))

mutate(tb,x=str_sub(species,1,3))
mutate(tb,x=str_replace(species,'e','-'))
mutate(tb,x=str_replace_all(species,'e','-'))
mutate(tb,x=str_remove(species,'e'))
mutate(tb,x=str_remove_all(species,'e'))

mutate(tb,x=str_c(' ',species,'  ',island,' '),len1=str_length(x)) |> 
  mutate(xx=str_trim(x),len2=str_length(xx))
mutate(tb,x=row_number()) |> 
  mutate(xx=str_pad(x,3,'left','0'))
```


## combine table vertically
```{r}
tb1=filter(tb0,sex=='male')
tb2=filter(tb0,sex=='female')
tb3=filter(tb0,is.na(sex))
bind_rows(tb1,tb2,tb3)
```

## combine table horizontally
```{r}
tb=mutate(tb0,id=row_number()) |> 
  select(id,everything())

tb1=select(tb,id,!is.numeric) 
tb2=select(tb,id,is.numeric) 

bind_cols(tb1,tb2)

tb1=slice_sample(tb1,n=100)
tb2=slice_sample(tb2,n=100)

inner_join(tb1,tb2,join_by(id==id))

left_join(tb1,tb2,join_by(id==id))

full_join(tb1,tb2,join_by(id==id))
```


## model matrix for multi regression
```{r}
tb=drop_na(tb0)
mx=model.matrix(~species+sex+bill_length_mm,tb)
mx[sample(1:nrow(mx),10),]
```


## use sqlite
```
library(RSQLite)
db='path/output0.sqlite'
con=dbConnect(SQLite(),db)
tbl='table0'

dbWriteTable(con,tbl,tb,overwrite=T)

rst=dbGetQuery(con,paste0("SELECT * FROM ",tbl,";"))
print(rst)


dbExecute(con,"
CREATE TABLE IF NOT EXISTS users (
    id INTEGER PRIMARY KEY,
    name TEXT,
    age INTEGER;
)")

dbExecute(con, "
INSERT INTO users (name, age)
  VALUES ('name1',10),('name2',20),('name3',30)
")

rst=dbGetQuery(con, "SELECT * FROM users;")
print(rst)

dbDisconnect(con)
```



## use parquet with nanoparquet library
```
library(nanoparquet)
df=read_parquet('data.parquet')

write_parquet(df,'data.parquet')
```



## input from web API
```{r}
#library(httr)
url0='https://api.excelapi.org/wikipedia/summary2?word='
word='api'
url=paste0(url0,word,'&lang=en','&mode=simple')

GET(url) |> content('text')


URLencode('ヤクルトスワローズ')
# "%E3%83%A4%E3%82%AF%E3%83%AB%E3%83%88%E3%82%B9%E3%83%AF%E3%83%AD%E3%83%BC%E3%82%BA"

enc=URLencode('ヤクルトスワローズ')
url=paste0(url0,enc,'&lang=jp','&mode=full')

GET(url) |> content('text',encoding='UTF-8')


GET('https://geoapi.heartrails.com/api/json?method=getPrefectures')
GET('https://geoapi.heartrails.com/api/json?method=getCities&prefecture=北海道')
GET('https://geoapi.heartrails.com/api/json?method=getTowns&prefecture=北海道&city=札幌市')

arg0=list(method='getPrefectures')
GET('https://geoapi.heartrails.com/api/json', query=arg0)

arg0=list(method='getCities',prefecture='北海道')
GET('https://geoapi.heartrails.com/api/json', query=arg0)


# error handling
url='https://geoapi.heartrails.com/api/json?method=getPrefectures'
res=GET(url)

if (status_code(res)==200) {
  obj=fromJSON(content(res,'text',encoding='UTF-8')) |> print()
} else {
  print(paste("Error:",status_code(res)))
}
```



## processing json
```{r}
#library(jsonlite)
fromJSON('[1,2,3]') # array -> vector

fromJSON('{"x":10,"y":20}') # object -> list of atomic
fromJSON('{"x":10,"y":20}') |> unlist() # named vector

fromJSON('[[11,12,13],[21,22,23],[31,32,33]]') # array of array -> matrix

fromJSON('[{"x":11,"y":12},{"x":21,"y":22}]') # array of object -> data.frame
fromJSON('[{"x":11,"y":12},{"x":21,"y":22}]') |> unlist() # index suffix named vector

fromJSON('{"x":[11,12,13],"y":[21,22,23]}') # object of arrays -> list of vectors
fromJSON('{"x":[11,12,13],"y":[21,22,23]}') |> unlist() # index suffix named vector

fromJSON('{"a":{"x":11,"y":12},"b":{"x":21,"y":22}}') # nested list
fromJSON('{"a":{"x":11,"y":12},"b":{"x":21,"y":22}}') |> unlist() # prefix named vector



json='{"glossary": {"title": "example glossary","GlossDiv": {"title": "S","GlossList": {"GlossEntry": {"ID": "SGML","SortAs": "SGML","GlossTerm": "Standard Generalized Markup Language","Acronym": "SGML","Abbrev": "ISO 8879:1986","GlossDef": {"para": "A meta-markup language, used to create markup languages such as DocBook.","GlossSeeAlso": ["GML", "XML"]},"GlossSee": "markup"}}}}}'

prettify(json)
minify(json)

obj=fromJSON(json,flatten=T)
str(obj) # list with list, not include array (a key has just one value)
v=unlist(obj) # vector of each element with json path name 
tibble(path=names(v),value=v)


json='[{"person":"Taro Yamada","pets":{"exists":true,"details":[{"name":"mike","kind":"cat"},{"name":"pochi","kind":"dog"}]}},{"person":"Jiro Suzuki","pets":{"exists":true,"details":[{"name":"koro","kind":"dog"}]}},{"person":"Sabro Tanaka","pets":{"exists":false}}]'

prettify(json)

obj=fromJSON(json,flatten=T)
str(obj) # data.frame with some list, include array (a key has plural value)
obj[1,]

tb=as_tibble(obj)
select(tb,!is.list) |> names() # json path has array
select(tb,is.list) |> names() # json path has list

tb0=tibble()
for(i in 1:nrow(tb)){
tb1=bind_cols(tb[i,]['person'], # for array ['json-path']
          tb[i,]['pets.exists'], # for array ['json-path']
          tb[i,]$pets.details) # for list $json-path
tb0=bind_rows(tb0,tb1)
}
tb0
```


## treat JSON response
```{r}
fromJSON('https://geoapi.heartrails.com/api/json?method=getPrefectures') # list has a vector
fromJSON('https://geoapi.heartrails.com/api/json?method=getCities&prefecture=北海道') |> glimpse() #list has a data.frama
fromJSON('https://geoapi.heartrails.com/api/json?method=getTowns&prefecture=北海道&city=札幌市') |> glimpse() # list has a data.frame

read_json('https://jsonplaceholder.typicode.com/users') |> glimpse() # list
fromJSON('https://jsonplaceholder.typicode.com/users',flatten = T) # simple data.frame


# https://zenn.dev/ttskch/articles/309423d26a1aaa
# https://jp-postal-code-api.ttskch.com/api/v1/6180000.json

url0='https://jp-postal-code-api.ttskch.com/api/v1/'
pcode='6180000'
url=paste0(url0,pcode,'.json')

obj=read_json(url)
str(obj) # list with list

obj=fromJSON(url,flatten=T) 
str(obj) # list with a simple data.frame

tibble(obj$postalCode,obj$address)



url0='https://query2.finance.yahoo.com/v8/finance/chart/'
code='4507.T'
range='3mo'
int='1d'
url=paste0(url0,code,'?range=',range,'&interval=',int)
# https://query2.finance.yahoo.com/v8/finance/chart/",$code$,".T?range=3mo&interval=1d

obj=read_json(url)
#str(obj)
glimpse(obj)

obj=fromJSON(url,flatten=T)
str(obj)
obj=obj$chart$result
str(obj$indicators.quote)
tibble(obj$timestamp[[1]],
       obj$indicators.quote[[1]]$open[[1]],
       obj$indicators.quote[[1]]$close[[1]],
       obj$indicators.quote[[1]]$low[[1]],
       obj$indicators.quote[[1]]$high[[1]],
       obj$indicators.quote[[1]]$volume[[1]],
       obj$indicators.adjclose[[1]]$adjclose[[1]])



obj=fromJSON('http://radioactivity.nsr.go.jp/data/ja/real/area_24000/2401_trend.json',flatten=T)
str(obj) # list with array has same number of elements

tibble(obj$day$airF$value,obj$day$est$value)
```



```
library(httr2)

url='https://api.example.com/data'

# make request object and excute
res=request(url) |> 
  req_perform()

# error handling
if (resp_status(res)==200) {
  data=resp_body_json(res)
  print(data)
} else {
  print(paste("Error:", resp_status(res)))
}



# httr2 can use OAuth2.0 and Bearer token 

    OAuth 2.0 Authentication

OAuth 2.0 is a protocol for performing authentication and authorization without passing the user's password directly to the API. A typical OAuth 2.0 flow is as follows:

    Client (Application) Registration: Register the application with an API provider (such as Google, Facebook) to obtain a client ID and client secret.
    Authorization Code Acquisition: When the user grants access to the application, an authorization code is issued.
    Access Token Acquisition: Use the authorization code and client secret to request and acquire an access token.
    Access Token Usage: Use the access token to access the API and make requests on behalf of the user.

    Bearer Token Authentication

Bearer token authentication is a simple method for accessing resources using an access token. This token is included in the HTTP request header, allowing the server to verify access permissions. It is often used with access tokens acquired through OAuth 2.0, and is used as follows:

Authorization: Bearer YOUR_ACCESS_TOKEN

    Example of OAuth 2.0 and Bearer Token with httr2

Below is an example of implementing the OAuth 2.0 flow using httr2. Here, we are using the client_credentials grant type, a simplified flow for obtaining a token with only a client ID and secret.
Preparation

    Installing the httr2 package:

install.packages("httr2")
library(httr2)

Obtaining an Access Token with OAuth 2.0 and Making an API Request

# Set the API endpoint and client information
auth_url <- "https://api.example.com/oauth2/token"
client_id <- "YOUR_CLIENT_ID"
client_secret <- "YOUR_CLIENT_SECRET"

# Create a request to obtain the access token
token_response <- request(auth_url) %>%
  req_body_form(
    grant_type = "client_credentials",
    client_id = client_id,
    client_secret = client_secret
  ) %>%
  req_perform()

# Extract the access token from the response
access_token <- resp_body_json(token_response)$access_token

# Make an API request using the access token
api_url <- "https://api.example.com/protected_resource"
response <- request(api_url) %>%
  req_auth_bearer_token(access_token) %>%
  req_perform()

# Display the result
data <- resp_body_json(response)
print(data)

Explanation

    Obtaining the Access Token:
        req_body_form is used to set the parameters required for the OAuth 2.0 client credentials flow (grant_type, client_id, client_secret).
        req_perform executes the request to obtain the access_token.

    Making an API Request Using the Bearer Token:
        req_auth_bearer_token is used to set the previously obtained access token.
        req_perform executes the request to access the protected resource.

Notes

    Since access tokens may expire, it is necessary to update the token periodically or use a refresh token.
    Each API provider may have specific differences in the OAuth 2.0 flow, so be sure to check the documentation.

With this code, you can easily access APIs protected by OAuth 2.0.
```


