---
title: "Various regression models"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
#date: "`r Sys.Date()`"
---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=T,warning=F,message=F,comment='')

suppressWarnings(
  suppressMessages(
    suppressPackageStartupMessages({
      library(stats)
      library(MASS)
      library(tidyverse)
      library(magrittr)
      
      library(palmerpenguins)
      
      library(lme4)
      library(rqlm) # coeff function
      library(lavaan)
      library(semPlot)
    })
  )
)
options(scipen=100,digits=3)
```

```{r}
tb0=penguins
tb=tibble(y=tb0$body_mass_g,
          x1=tb0$flipper_length_mm,
          x2=tb0$bill_length_mm,
          x3=tb0$bill_depth_mm,
          c1=factor(tb0$species),
          c2=factor(tb0$sex)) |> 
  drop_na()
```




# Linear models
```{r}
fitrst1=function(fit){
  print(coef(fit))
  print(confint(fit))
  predict(fit,tb) |> head() |> print()
  print(cor(tb$y,predict(fit,tb)))
}
```
\
$y=b0+b1*x1$
```{r}
f='y~x1'
fit=lm(f,tb)
fitrst1(fit)

xx=tibble(x1=seq(min(tb$x1),max(tb$x1),length.out=7))
bind_cols(xx,predict(fit,xx,interval='confidence'))
bind_cols(xx,predict(fit,xx,interval='prediction'))
```
\
$y=b1*x$
```{r}
f='y~0+x1'
fit=lm(f,tb)
fitrst1(fit)
```
\
$y=b0+b1*x1+b2*x2$
```{r}
f='y~x1+x2'
fit=lm(f,tb)
fitrst1(fit)
```


### all explanatory variables
```{r}
f='y~.'
fit=lm(f,tb)
fitrst1(fit)
```


### exclude some explanatory variables
```{r}
f='y~.-c1-c2'
fit=lm(f,tb)
fitrst1(fit)
```
\
$y=b0+b1(x1+x2)$
```{r}
f='y~I(x1+x2)'
fit=lm(f,tb)
fitrst1(fit)
```
\
$y=b0+b1*x1+b2*x1^2$
```{r}
f='y~x1+I(x1^2)'
fit=lm(f,tb)
fitrst1(fit)
```
\
$y=b0+b1(x1*x2)$
```{r}
f='y~x1:x2'
fit=lm(f,tb)
fitrst1(fit)
```
\
$y=b0+b1*x1+b2*x2+b12*x1x2$
```{r}
f='y~x1*x2'
fit=lm(f,tb)
fitrst1(fit)
```
\
$y=b0+b1*x1+b2*x2+b3*x3+b12*x1x2+b13*x1x3+b23*x2x3$
```{r}
f='y~(x1+x2+x3)^2'
fit=lm(f,tb)
fitrst1(fit)
```

\

# Random effect of group c[1,l] r_0c, r_1c
```{r}
#library(lme4) #lmer function, glmer function
#library(rqlm)

fitrst2=function(fit){
  print(fit)
  print(coeff(fit))
  print(ranef(fit))
  predict(fit,tb) |> head() |> print()
  print(cor(tb$y,predict(fit,tb)))
}

cat('c1: ',levels(tb$c1),'\n')
cat('c2: ',levels(tb$c2),'\n')
```

## random intercept
$y=b0+r_{0c}$
```{r}
f='y~(1|c1)'
fit=lmer(f,tb)
fitrst2(fit)
```
\
$y=b0+r_{0c1}+r_{0c2}$
```{r}
f='y~(1|c1)+(1|c2)'
fit=lmer(f,tb)
fitrst2(fit)
```
\
$y=b0+r_{0c1}+r_{0c2,c1}$
```{r}
f='y~(1|c1)+(1|c1:c2)'
fit=lmer(f,tb)
fitrst2(fit)
```



## random intercept, ANOVA model
$y=b0+r_{0c1}+b1*x1$
```{r}
f='y~(1|c1)+x1'
fit=lmer(f,tb)
fitrst2(fit)
```


## random slope
$y=b0+(b1+r_{1c})x$\
intercept and slope is uncorrelated by group
```{r}
f='y~x1+(x1||c1)'
fit=lmer(f,tb)
fitrst2(fit)
```


## random mixed
$y=b0+r_{0c}+(b1+r_{1c})x$\
intercept and slope is correlated by group
```{r}
f='y~x1+(x1|c1)'
fit=lmer(f,tb)
fitrst2(fit)

xx=tibble(x1=seq(min(tb$x1),max(tb$x1),length.out=21),
          c1=factor(rep(c('Adelie','Gentoo','Chinstrap'),7)))
predict(fit,xx)
```

\

# Generalized linear models
```{r}
f='y~x1'
```

## normal regression 
```{r}
fit=glm(f,tb,family=gaussian(link='identity'))
fit
coef(fit)
confint(fit)
predict(fit,tb) |> head()
print(cor(tb$y,predict(fit,tb)))

xx=tibble(x1=seq(min(tb$x1),max(tb$x1),length.out=7))
predict(fit,xx)
```

## normal log regression
```{r}
fit=glm(f,tb,family=gaussian(link='log'))
fit
coef(fit)
confint(fit)
predict(fit,tb) |> head() |> exp()
print(cor(tb$y,exp(predict(fit,tb))))


xx=tibble(x1=seq(min(tb$x1),max(tb$x1),length.out=7))
predict(fit,xx) |> exp()
```

## poisson regression
```{r}
fit=glm(f,tb,family=poisson(link='log'))
fit
coef(fit)
confint(fit)
predict(fit,tb) |> head() |> exp()
print(cor(tb$y,exp(predict(fit,tb))))

xx=tibble(x1=seq(min(tb$x1),max(tb$x1),length.out=7))
predict(fit,xx) |> exp()
```

\

# Generalized linear mixed models
```{r}
f='y~x1+(x1|c1)'
```

## normal mixed regression
```{r}
fit=glmer(f,tb,family=gaussian('identity'))
fit
coeff(fit)
ranef(fit)
coef(fit)
predict(fit,tb) |> head()
print(cor(tb$y,predict(fit,tb)))

xx=tibble(x1=seq(min(tb$x1),max(tb$x1),length.out=21),
          c1=factor(rep(c('Adelie','Gentoo','Chinstrap'),7)))
predict(fit,xx)
```

## normal log mixed regression
```{r}
fit=glmer(f,tb,family=gaussian('log'))
fit
coeff(fit)
ranef(fit)
coef(fit)
predict(fit,tb) |> head() |> exp()
print(cor(tb$y,exp(predict(fit,tb))))

xx=tibble(x1=seq(min(tb$x1),max(tb$x1),length.out=21),
          c1=factor(rep(c('Adelie','Gentoo','Chinstrap'),7)))
predict(fit,xx) |> exp()
```

## possion mixed regression
```{r}
fit=glmer(f,tb,family=poisson('log'))
fit
coeff(fit)
ranef(fit)
coef(fit)
predict(fit,tb) |> head() |> exp()
print(cor(tb$y,exp(predict(fit,tb))))

xx=tibble(x1=seq(min(tb$x1),max(tb$x1),length.out=21),
          c1=factor(rep(c('Adelie','Gentoo','Chinstrap'),7)))
predict(fit,xx) |> exp()
```

\

# Strucutured Equation Model, hierarchical model has latent variable
```{r}
#library(lavaan)
#library(semPlot)

fitsem=function(fit){
  print(fitMeasures(fit, "df"))
  print(standardizedSolution(fit))
  semPaths(fit, what='std', residuals=F,
          edge.label.cex=1, layout='tree') # layout: circle,spring
}

n=20

tb0=tibble(x=rnorm(n,0,1),y=rnorm(n,x,1)) |>
  scale()
mdl='y~x'
fit=sem(mdl,tb0,std.lv=T)
fitsem(fit)


tb0=tibble(x1=rnorm(n,0,1),x2=rnorm(n,x1,1),y=rnorm(n,x1+x2,1)) |> 
  scale()
mdl='y~x1+x2'
fit=sem(mdl,tb0,std.lv=T)
fitsem(fit)


tb0=tibble(x1=rnorm(n,0,1),x2=rnorm(n,x1,1),y=rnorm(n,x2,1)) |> 
  scale()
mdl='
  x2~x1
  y~x2'
fit=sem(mdl,tb0,std.lv=T)
fitsem(fit)


tb0=tibble(x1=rnorm(n,0,1),x2=rnorm(n,x1,1),x3=rnorm(n,x2,1),y=rnorm(n,x3,1)) |> 
  scale()
mdl='
  x2~x1
  x3~x2
  y~x3'
fit=sem(mdl,tb0,std.lv=T)
fitsem(fit)


tb0=tibble(x1=rnorm(n,0,1),x2=rnorm(n,x1,1),x3=rnorm(n,0,1),y=rnorm(n,x2+x3*2,1)) |> 
  scale()
mdl='
  x2~x1
  y~x2+x3
  x2~~0*x3'
fit=sem(mdl,tb0,std.lv=T)
fitsem(fit)


# apply latent variable to model
# to specify latent factor, do factor analysis before SEM

set.seed(1)

x1=rnorm(n,0,1) #explanatory
x2=rnorm(n,0,1) #explanatory
x3=rnorm(n,0,1) #explanatpry
xl1=rnorm(n,x1+x2,1) #latent
xl2=rnorm(n,x1+x3,1) #latent
x4=rnorm(n,xl1+xl2*2,1) #objective
x5=rnorm(n,xl1*2+xl2,1) #objective

set.seed(Sys.time())

tb0=tibble(x1,x2,x3,x4,x5) |> 
  scale() |> 
  as_tibble()

# factor analysis
# decide a number of factor
eigen(cor(tb0))$values |> print() |> # see a number of eigen >1
  plot(type='b') # screw plot

factanal(tb0,factors=2,rotation='promax') # see relation between each variable and latent factor 
cor(tb0) # see low correlation to reject covariance on model

mdl='
  l1=~x1+x3 # latent =~ observed explanatory
  l2=~x2+x3
  x4~l1 # observed objective ~ latent
  x5~l1+l2
  x1~~0*x2 # reject covariance
  l1~~0*l2'
fit=sem(mdl,tb0,std.lv=T)
fitsem(fit)
```


```{r}
tb1=bind_cols(select(tb,is.numeric) |> scale(), select(tb,!is.numeric))
tb1=model.matrix(~.,tb1)
tb1=tb1[,-1]
head(tb1)
cor(tb1)

mdl='y~c2male' # SEM can apply binary variable
fit=sem(mdl,tb1,std.lv=T)
fitsem(fit)

mdl='y~x1+c1Chinstrap+c1Gentoo'
fit=sem(mdl,tb1,std.lv=T)
fitsem(fit)

mdl='y~x1+c1Chinstrap+c1Gentoo+c2male'
fit=sem(mdl,tb1,std.lv=T)
fitsem(fit)

mdl='
  x2~c1Chinstrap+c1Gentoo+c2male
  x3~c1Chinstrap+c1Gentoo+c2male'
fit=sem(mdl,tb1,std.lv=T)
fitsem(fit)


eigen(cor(tb1))$values |> print() |> # see a number of eigen >1
  plot(type='b') # screw plot


# 2 factors model
factanal(tb1,factors=2,rotation='promax') 

mdl='
  l1=~c1Chinstrap+c1Gentoo+c2male
  l2=~c1Gentoo+c2male
  y~l1+l2
  x1~l1+l2
  x2~l2
  x3~l1
  l1~~l2
  c1Chinstrap~~0*c2male
  c1Gentoo~~0*c2male
  x2~~0*x3'
fit=sem(mdl,tb1,std.lv=T)
fitsem(fit)


# 3 factors model
factanal(tb1,factors=3,rotation='promax') 

mdl='
  l1=~c1Chinstrap+c1Gentoo
  l2=~c1Chinstrap
  l3=~c2male
  y~l1+l3
  x1~l1
  x2~l2+l3
  x3~l1+l3
  l2~~l3
  l1~~0*l2
  l1~~0*l3
  l2~~0*l3
  x2~~0*x3'
fit=sem(mdl,tb1,std.lv=T)
fitsem(fit)
```

